Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-12 appendix.tex
rough revision done

\subsubsection{Stability without vertical shear}
In the absence of vertical shear $q=0$, Eq. \ref{simple_growth} can be
written as \begin{align}
  \hat{\sigma}^2\khat^2 =
  \left(L-\hat{\sigma}^2\right)\left(1-\hat{\sigma}^2\right). \label{iso_disp_full}
\end{align}
This is just the dispersion relation for axisymmetric waves in a
vertically isothermal disk already discussed in previous works 
\citep[e.g.][]{okazaki87,takeuchi98,tanaka02,zhang06,ogilvie13,barker14}. In
this case the solutions are Hermite polynomials, $W\propto
\He_L(\zhat)$.  The eigenfrequency $\hat{\sigma}$ is real so the disk
is stable, consistent with \S\ref{integral_relation}. The
low-frequency solution to Eq. \ref{iso_disp_full} correspond to
inertial waves \citep{balbus03}.

>>>>

\subsubsection{Stability without vertical shear}\label{stable_novshear}
In the absence of vertical shear $q=0$, Eq. \ref{simple_growth} can be
written as \begin{align}
  \hat{\sigma}^2\khat^2 =
  \left(L-\hat{\sigma}^2\right)\left(1-\hat{\sigma}^2\right). \label{iso_disp_full}
\end{align}
This is just the dispersion relation for axisymmetric waves in a
vertically isothermal disk already discussed in previous works 
\citep[e.g.][]{okazaki87,takeuchi98,tanaka02,zhang06,ogilvie13,barker14}. In
this case the solutions are Hermite polynomials, $W\propto
\He_L(\zhat)$.  The eigenfrequency $\hat{\sigma} = \hat{\omega}$ is real so the disk
is stable, consistent with \S\ref{integral_relation}. The
low-frequency solution to Eq. \ref{iso_disp_full} correspond to
inertial waves \citep{balbus03}. In the low frequency limit
$|\hat{\omega}|\ll 1$ and assuming $L\geq 1 $, the 
dispersion relation is $\hat{\omega}^2\khat^2 = L$, or
$\hat{\omega}\propto \khat^{-1}$ for fixed $L$. (Low frequency thus 
require $\khat\gg \sqrt{L}$.) This inverse relation has been qualitatively
observed in numerical simulations of \cite{stoll14}.

### 2015-03-12 linear_adia.tex
updating results
rough revision done

Eq. \ref{nearly_iso_explicit} becomes 
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
which requires
\begin{align}
  B(\hat{\sigma}) + \alpha(\hat{\sigma}) =
  M\left[A-2\alpha(\hat{\sigma})\right].\label{adia_disp0} 
\end{align}
% We remark that for $q=0$ and $\beta\to\infty$, Eq. \ref{adia_disp0} is
% the same as Eq. 54 in \cite{lubow93} when the low-frequency, Keplerian
% approximation is made in the latter. 
% Eq. \ref{nearly_iso_explicit} is same form
% as the governing equation for the adiabatic case
% (Eq. \ref{adia_iso4}). However, the coefficients $\widetilde{B}$ and $\widetilde{C}$
% have a complicated dependence on $\hat{\sigma}$. 
% Applying the same solution method as in \S\ref{adia_explicit}, we 
% find the adiabatic dispersion relation, 
This is a $6^\mathrm{th}$ order 
polynomial in $\hat{\sigma}$:  
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
The dispersion relation $\hat{\sigma}=\hat{\sigma}(\khat)$ generally
requires a numerical solution. However, for the fundamental mode
($M=0$) some analytic progress can be made, which we consider next.

@@

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI mode for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
$\khat\gg1$, we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3 \notag\\
    &+ 2\beta\gamma(\hat{\omega}\khat)^4.  
\label{relax_cond2}
\end{align}
This is a pair of simultaneous equations for
$\hat{\omega}$ and $\beta$. In Appendix \ref {iso_stable} we discuss
stable solutions without vertical shear which yield
$\hat{\omega}\propto \khat^{-1}$ for $\khat\gg 1$
(Eq. \ref{iso_stable_disp}).  We are thus motivated to seek solutions
to Eq. \ref{relax_cond1}---\ref{relax_cond2} with  
% we wish to seek a criterion independent of k 
% \begin{align}
%   \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
%\end{align}
\begin{align}
  \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
  |\hat{\omega}\khat|=O(1). 
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}\label{beta_crit0}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$.

@@

% In Fig. \ref{relax_bcrit} we plot the critical thermal relaxation
% timescale obtained from solving Eq. \ref{relax_disp_fund} numerically 
% for a disk with $\epsilon = 0.05$, $q=-1$ and $\gamma=1.4$, for which
% we have $\beta_\mathrm{crit}=0.125$ from Eq. \ref{iso_vsi_cond}. Our
% upper estimate agrees well with the numerical solution for $\khat\gtrsim
% 10$.

% \begin{figure}
%   \includegraphics[width=\linewidth]{figures/bcrit_theory}
%   \caption{Condition on the thermal relaxation timescale $\beta$ for
%     the existence of unstable fundamental VSI mode in a vertically
%     isothermal disk with $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$.  
%     The dotted line denotes the asymptotic value
%     $\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$ for large $\khat$. 
%     \label{relax_bcrit}}  
% \end{figure}

% \subsubsection{Lower limit on the thermal relaxation timescale for the
%   fundamental adiabatic VSI mode}
% {\bf Note: for reference only. Not to be included in paper.} 
% Fig. \ref{relax_bcrit} indicates there exists a lower limit on the
% thermal relaxation time for the system to behave adiabatically and
% support the fundamental adiabatic VSI. This critical
% timescale satisfies Eq. \ref{relax_cond1}---\ref{relax_cond2} with 
% \begin{align}
%   \beta \sim \beta_0\khat^{1/2},\quad \hat{\omega}  \sim
%   \hat{\omega}_0\khat^{-1/2} \text{ as } \khat\to\infty.
% \end{align}
% The constants $\beta_0$ and $\hat{\omega}_0$ are determined by the
% simultaneous equations
% \begin{align}
%   & 0 = \gamma\beta_0^2(1-\gamma) + 2\epsilon q \gamma
%   \beta_0\hat{\omega}_0 - \hat{\omega}_0^2 +
%   \beta_0^2\gamma^2\hat{\omega}_0^4.\\
%   & 0= \beta_0(1-\gamma) + \epsilon q \hat{\omega}_0 - \epsilon q
%   \gamma^2 \beta_0^2\hat{\omega}_0^3. 
% \end{align}
% For $\epsilon = 0.05$, $q=-1$ and $\gamma=1.4$, the appropriate
% solution is $\beta_0 = 10.33$ and $\hat{\omega}_0 = 0.74$.

% If we follow through the solution method as before, we obtain a
% dispersion relation 
% \begin{align}
%   &\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) \notag\\ &=
%   \left(\frac{1}{\gamma}-\frac{1}{2}\right) +\frac{\ii}{2}\epsilon q
%   \hat{k} + \frac{1}{2}\left(2M+1\right)\left(A^2 + 4C\right)^{1/2}\label{adia_disp}.
% \end{align}
% This equation is the same as Eq. 54 in \cite{lubow93} when $q=0$ and
% the low-frequency, Keplerian approximation is made for the latter.

% Note that when $\gamma>1$ and $q=0$, the term with $\ln\rho^{\prime
%   2}$ in the integrand (third term on the RHS) represents
% stabilization by the vertical entropy gradient (real buoyancy frequency). 
% It gives an imaginary contribution to $\hat{\sigma}$ when $q\neq0$,
% but this is a factor $|\epsilon/\hat{k}|$ smaller than the real
% (stabilizing) contribution, 
% since $\epsilon\ll 1$ for a thin disk and we (usually) consider
% $\hat{k}\gg1$. Thus, instability is expected to be associated with the 
% last term on the RHS of Eq. \ref{adia_integral}.

% \subsubsection{Explicit solutions in the thin-disk limit}\label{adia_explicit}
% We can construct solutions in the thin-disk limit, in which 
% $\ln(\rho/\rho_0) \simeq -\zhat^2/2$. Eq. \ref{adia_iso3} becomes   
% \begin{align}
%   0 = \delta v_z^{\prime\prime} - \zhat A \delta v_z^\prime + \left(B
%     - C \zhat^2\right)\delta v_z, \label{adia_iso4} 
% \end{align}
% with
% \begin{align}
%   &A \equiv 1 + \ii \epsilon q \hat{k},\\
%   &B \equiv \hat{\sigma}^2\left(\frac{1}{\gamma} + \hat{k}^2\right) -
%   \left(\frac{1}{\gamma} + \ii \hat{k} \epsilon q\right)\\
%   &C \equiv \frac{\left(\gamma-1\right)}{\gamma}\hat{k}^2\left(1 - \frac{\ii
%       \epsilon q}{\hat{k}}\right).\label{adia_thin}
% \end{align}

% Inserting the definition of $B$ we get
% \begin{align}
%   &\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) \notag\\ &=
%   \left(\frac{1}{\gamma}-\frac{1}{2}\right) +\frac{\ii}{2}\epsilon q
%   \hat{k} + \frac{1}{2}\left(2M+1\right)\left(A^2 + 4C\right)^{1/2}\label{adia_disp}. 
% \end{align}
% This gives the eigenfrequency $\hat{\sigma}$ and completes the solution
% description.

% %Recalling that $B$ depends $\hat{\sigma}$, Eq. \ref{adia_disp} gives
% %the eigenfrequency as a function of disk paramters.

% \subsubsection{The case of $\gamma=2$}
% In principle, we can solve Eq. \ref{adia_disp} for the mode
% frequencies and growth rates as a function of the disk parameters and
% the radial wavenumber. This is complicated in 
% general, but simplifies for $\gamma=2$, for which the quantity
% $A^2+4C$ is real. In this case we find
% \begin{align}
%   \hat{\nu}^2 =
%   \frac{1}{2}\frac{(2M+1)}{\sqrt{1+2\hat{k}^2}}&\left[
%     \sqrt{1-\frac{4M(M+1)\hat{k}^2\epsilon^2q^2}{(2M+1)^2(1+2\hat{k}^2)}}\right.\notag\\
%   &\phantom{=}
%   \left.- \sqrt{1 - \frac{\hat{k}^2\epsilon^2q^2}{(1+2\hat{k}^2)}}\right].
% \end{align}
% In the limit of a thin disk $\epsilon\to0$ or weak shear $|q|\to0$, we
% have 
% \begin{align}
%   \hat{\nu}^2 = \frac{\epsilon^2
%     q^2}{4(2M+1)}\frac{\hat{k}^2}{(1+2\hat{k}^2)^{3/2}},\label{gam2_growth}
% \end{align}
% and the maximum growth rate occurs at $\hat{k}=\hat{k}_\mathrm{opt}=1$,  
% \begin{align}
%   \mathrm{max}(\hat{\nu}) &= \frac{|\epsilon
%     q|}{2(3^{3/4}\sqrt{2M+1})}\label{gam2_max_growth}.
%   % & \leq \frac{|\epsilon
%   % q|}{2\times3^{3/4}}\notag.
% \end{align}
% Unlike isothermal perturbations, Eq. \ref{gam2_max_growth} shows that
% maximum growth rate occurs at $M=0$, and growth rates decrease with increasing $M$. This
% is despite the fact that in the thin-disk approximation 
% $d\Omega^2/dz\propto z$, so there is infinite vertical shear
% available.

% The acceptable solution for $\alpha$ when $\gamma=2$ is
% \begin{align}
%   \alpha = \frac{1}{2}\left[\left(1+\ii\epsilon q \hat{k}\right) -
%     \sqrt{1 + \hat{k}^2\left(2-\epsilon^2 q^2\right)}\right].\label{gam2_alpha}
% \end{align}
% Since $\epsilon\ll 1$, $\real\alpha<0$, so the vertical velocity
% eventually decays for large heights. For $\hat{k}\gg 1$ the
% perturbation decays rapidly away from the midplane. This is again
% unlike isothermal perturbations, whose magnitude increase with $|z|$.

% %gauss decay means boundary conditions have no influence, all perts go
% %to zero

% \subsubsection{Comparison with $\gamma=1$}\label{adia_compare_gam1}
% We can compare growth rates obtained for $\gamma=2$ to those for
% $\gamma=1$ in the previous section. If we compare growth rates at their 
% respective optimum radial wavelengths, we find
% that for $\gamma=2$ the \emph{maximum} growth rate, occuring
% at $M=0$, is still $\sim2$ times smaller than the \emph{minimum} growth
% rate for $\gamma=1$. That is,
% \begin{align}
%   &\mathrm{max}(\hat{\nu}) \geq \frac{|\epsilon q|}{2\sqrt{1+|\epsilon
%       q|}} & \text{at }\hat{k} = \hat{k}_\mathrm{opt} \text{ for }   \gamma=1,  \\
%   &\mathrm{max}(\hat{\nu}) \leq \frac{|\epsilon q|}{2\times 3^{3/4}}
%            & \text{at }\hat{k} = \hat{k}_\mathrm{opt} \text{ for } \gamma=2.
% \end{align}

% We can also compare growth rates in the limit $\hat{k}\gg1$ using
% Eq. \ref{simple_growth} and Eq. \ref{gam2_growth},
% \begin{align}
%   &\hat{\nu} \geq \sqrt{\frac{|\epsilon q|}{2\hat{k}}}
%   &  \text{as }  \hat{k}\to\infty  \text{ for }  \gamma=1, \\
%   &\hat{\nu} \leq \frac{|\epsilon q|}{2^{7/4}\sqrt{\hat{k}}}
%   &\text{as }  \hat{k}\to\infty  \text{ for }  \gamma=2.\label{gam2_growth_rate}
% \end{align}
% Then the growth rate of perturbations with large $|\hat{k}|$ in a
% $\gamma=2$ disk is at most $2^{-5/4}\sqrt{|\epsilon q|}$ times that in a
% $\gamma=1$ disk. For $\epsilon = 0.1$ and $|q|=1$, this factor is
% $\sim 0.1$.

% In Fig. \ref{adia_growth} we plot the eigenfrequencies as a
% function of the radial wavenumber $\hat{k}$, for a range of adiabatic
% indices, using Eq. \ref{adia_disp}. The low-frequency approximation
% is questionable for $\hat{k}\lesssim4$ since
% $|\hat{\sigma}|^2\gtrsim0.2$. For large $\hat{k}$ the presence of a
% positive vertical entropy gradient is strongly stabilizing. This is
% because the vertical shear grows as $z$ away from the midplane (and
% has a maximum when the thin-disk approximation is relaxed), but square
% of the bouyancy frequency, which is stabilizing, grows as $z^2$ away
% from the midplane. %The entropy $P/\rho^\gamma$ formally diverges   
% %for $z\to\infty$ even without the thin-disk approximation.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/rate_theory_om}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/rate_theory_nu}
%   \caption{Real frequency (top) and growth rate (bottom) of the fundamental VSI mode in vertically 
%     isothermal disks ($\Gamma=1$) subject to adiabatic perturbations,
%     calculated in thin disk approximation from Eq. \ref{adia_disp}
%     with $M=0$, $\epsilon=0.1$ and  
%     $|q|=1$. Frequencies are shown as a function of the radial
%     wavenumber for different values of the adiabatic index
%     $\gamma$. For $\gamma\equiv 1$ the perturbations are isothermal
%     and there is no vertical entropy gradient.\label{adia_growth}}  
% \end{figure}

>>>>

Eq. \ref{nearly_iso_explicit} becomes 
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
which requires
\begin{align}
  B(\hat{\sigma}) + \alpha(\hat{\sigma}) =
  M\left[A-2\alpha(\hat{\sigma})\right].\label{adia_disp0} 
\end{align}
% We remark that for $q=0$ and $\beta\to\infty$, Eq. \ref{adia_disp0} is
% the same as Eq. 54 in \cite{lubow93} when the low-frequency, Keplerian
% approximation is made in the latter. 
% Eq. \ref{nearly_iso_explicit} is same form
% as the governing equation for the adiabatic case
% (Eq. \ref{adia_iso4}). However, the coefficients $\widetilde{B}$ and $\widetilde{C}$
% have a complicated dependence on $\hat{\sigma}$. 
% Applying the same solution method as in \S\ref{adia_explicit}, we 
% find the adiabatic dispersion relation, 
For ease of analysis, we square Eq. \ref{adia_disp0} to obtain a 
a polynomial in $\hat{\sigma}$,  
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
The dispersion relation $\hat{\sigma}=\hat{\sigma}(\khat)$ generally
requires a numerical solution. However, for the fundamental mode
($M=0$) some analytic progress can be made, which we consider next.

@@

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI mode for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
$\khat\gg1$, we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3 \notag\\
    &+ 2\beta\gamma(\hat{\omega}\khat)^4.  
\label{relax_cond2}
\end{align}
This is a pair of simultaneous equations for
$\hat{\omega}$ and $\beta$. Now, low-frequency axisymmetric waves in
the absence of vertical shear are stable and  satisfy
$\hat{\omega}\propto \khat^{-1}$ (Appendix \ref
{stable_novshear}, Eq. \ref{iso_disp_full}).  We are thus motivated to
seek solutions to Eq. \ref{relax_cond1}---\ref{relax_cond2} with  
% we wish to seek a criterion independent of k 
% \begin{align}
%   \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
%\end{align}
\begin{align}
  \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
  |\hat{\omega}\khat|=O(1). 
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}\label{beta_crit0}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$.

@@
